---
title: "Lab | Introduction to R and Tidyverse"
output: html_notebook
---

```{r}
install.packages("tidyverse")
library(tidyverse)
```

# Guillermo's Notebook

# Task 1: Explore the Data

```{r}
employees <- data.frame(
		  ID = 1:6,
		  Name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
		  Age = c(25, 30, 35, 40, 45, 50),
		  Department = c("HR", "IT", "Finance", "IT", "HR", "Finance"),
		  Salary = c(50000, 60000, 70000, 80000, 55000, 75000)
		)
```

**1.** Print the employees data frame to the console.

```{r}
print(employees)

```

**2.** Use the str() function to inspect the structure of the data frame.

```{r}
str(employees)
```

**3.** Use the summary() function to get a summary of the data.

```{r}
summary(employees)
```

## Task 2: Basic Data Manipulation with dplyr

**1.** Filter rows: Create a new data frame that includes only employees who work in the "IT" department.

```{r}
filtered_employees <- employees %>% filter(Department == "IT")
print(filtered_employees)
```

**2.** Select columns: Create a new data frame that includes only the Name and Salary columns.

```{r}
new_df <- employees %>% select(Name, Salary)
print(new_df)
```

**3.**Add a new column: Add a new column called Bonus that calculates a 10% bonus for each employee based on their salary.

```{r}
updated_df = new_df %>% mutate(Bonus = Salary * 1.1 )
print(updated_df)
```

**4.**Sort rows: Sort the employees data frame by Salary in descending order.

```{r}
sorted_df <- employees %>% arrange(desc(Salary))
print(sorted_df)
```

**5.**Summarize data: Calculate the average salary for each department.

```{r}
av_sal_tot <- employees %>% summarize (mean(Salary))
print(av_sal_tot)

av_sal_dep <- employees %>% group_by(Department) %>% summarize(Average_Salary = mean(Salary, na.rm = TRUE))  # The na.rm = TRUE part ensures that any NA values are ignored when calculating the mean.
print(av_sal_dep)

```

## Optional Exercises

### **Extra 1**: Advanced Data Manipulation

**1.** Group and summarize: Group the data by Department and calculate the total salary expenditure for each department.

```{r}
sal_exp <- employees %>% group_by(Department) %>% summarize(dep_exp = sum(Salary, na.rm = TRUE))
print(sal_exp)

```

**2.** Filter and mutate: Create a new data frame that includes only employees older than 30 and adds a column called Experience that assumes each employee has Age - 25 years of experience.

```{r}
fil_mut_df <- employees %>% filter(Age > 30) %>% mutate(Experience = Age - 25)
print(fil_mut_df)
```

### Extra 2: Challenge

**1.** Combine operations: Create a new data frame that includes employees from the "HR" department, adds a Bonus column (10% of salary), and sorts the data by Bonus in descending order.

```{r}
hr_df <- employees %>% filter(Department == "HR") %>% mutate(Bonus = Salary * 1.1 ) %>% arrange(desc(Bonus))
print(hr_df)
```

**2.** Visualize data: Use ggplot2 to create a bar plot showing the total salary expenditure by department.

```{r}
# Create the bar plot
ggplot(sal_exp, aes(x = Department, y = dep_exp)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Total Salary Expenditure by Department",
       x = "Department",
       y = "Total Salary Expenditure") +
  theme_minimal()

```
